\section{2023-09-13}

\subsection{Intro}
Аннотации, аннотации, аннотации!!!

Примерно по тематике лекций. Найти статью, приближенную к теме лекции. Формат: А4. Плюсы и минусы аннотаций.

\begin{itemize}
	\item Lectures (2 times a week)
	\item Labworks (2 times a week)
	\item 
		\begin{itemize}
			\item Annotations
			\item 2 tests in CDO
			\item Exam
			\item Bonuses for detected mistakes \& inordinary solutions 
		\end{itemize}
\end{itemize} 

\begin{note}[]
	Main goal is \textbf{to learn a lot of information fastly and effectively}
\end{note}

\subsection{Information theory}

Classic definition:
\begin{definition}[Probability]
	$p(A) = \frac{m}{n}$
\end{definition}

Statistic definition:
\begin{definition}[]
	$p(A) = \lim_{n \to \infty} \frac{m}{n}$
\end{definition}

\begin{property}[]
	$0 \leq p(A) \leq 1$
\end{property}

\subsection{Hartley measure}

A system $S$ can be in $N$ conditions. Then, we can represent a condition using powers of 2. Because of  $P = \frac{1}{N}$, 

\subsection{Shannon's measure}

\[
	i(S) = -\sum_{i = 1}^{N} p_{i} \cdot \log_2 p_{i}
\]

$N$ -- amount of system's conditions

$p_i$ -- probability of $S$ in condition $i$ 

\begin{example}[]
	We have 3 jokers, 3 aces, 1 king, 1 queen, 1 jack. What's the amount of information?
\end{example}

\subsection{Why different volume on different devices?}

\begin{center}
	\begin{tabular}{|c|c|c|}
		 Приставки в СИ & Новые двоичные приставки  & $\Delta$\\
		$kB = 10^{3} byte$ & $KiB = 2^{10} byte$ & $2\%$ \\
		$MB = 1e 6$ & $MiB = 2e 20$ & $5\%$ \\
		$ \dots $ & $ \dots $ & $ \dots $ \\
	\end{tabular}
\end{center}

\subsection{Unary notation}

The main disadvantage -- absence of zero.

Let $X_{(q)} = x_{n-1}x_{n-2} \dots x_0. x_{-1}x_{-2} \dots x_{-m}$

\[
	X_{(q)} = \sum_{i = -m}^{n-1} x_i \cdot q^{i}
\]

\begin{problem}[]
	$0,8125_{10} = \ ?_{(2)}$
\end{problem}


\subsection{Optimal notation}
\begin{problem}[]
	Robinson Cruso has found 60 stones. How many days we cane code using this 60 stones?
\end{problem}

$e$ is optimal.

Optimal means best on some cryteria/option

Обеспечивать, число и функциональность

\subsection{Transposition to any notation}

\subsection{Bergman's Notation}
$\forall x \in \R: x = \sum_{k = -n}^{n-1} d_k \cdot z^{k}$, where $d_{k} \in \left\{ 0, 1 \right \}, z = \frac{1 + \sqrt{5}}{2} $

Slide is fucked up

\subsection{Zekendorf's notation}

$x = \sum_{k = 1}^{n} d_{k}F_{k}$, where $d_{k} \in \left\{ 0, 1 \right \} , F_{k}$ are Fibonnacci numbers

2 <<1's>> in a row is forbidden!

\subsection{Factorial notation}

$x = \sum_{k = 1}^{n} d_{k} k!$
